\documentclass{article}

\usepackage{arxiv}

% Strip the "A Preprint" header/footer that overlaps with long titles.
\renewcommand{\headeright}{}
\renewcommand{\undertitle}{}
\renewcommand{\shorttitle}{}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\cfoot{\thepage}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{cite}
\usepackage{microtype}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta}

\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  showstringspaces=false,
  breaklines=true,
  frame=single,
  rulecolor=\color{gray!30},
  language=Python,
  commentstyle=\color{gray},
  keywordstyle=\color{blue!70!black}\bfseries,
  stringstyle=\color{green!50!black},
}

\title{mcos-decoder: A Pure-Python Reader for MATLAB Computer Vision Toolbox \texttt{groundTruth} Objects}

\author{
  Burak Ozdemir \\
  \texttt{bozdemir@gmail.com} \\
  \href{https://github.com/bozdemir/mcos-decoder}{\texttt{github.com/bozdemir/mcos-decoder}} \\
  DOI: \href{https://doi.org/10.5281/zenodo.19728531}{\texttt{10.5281/zenodo.19728531}}
}

\begin{document}
\maketitle

\begin{abstract}
The MATLAB Computer Vision Toolbox stores video annotations as
\texttt{groundTruth} MCOS (MATLAB Class Object Storage) instances inside
\texttt{.mat} files. Standard Python loaders (\texttt{scipy.io.loadmat},
\texttt{pymatreader}, \texttt{mat73}) surface this data only as opaque
\texttt{\_\_function\_workspace\_\_} blobs, blocking dataset users
without a MATLAB licence. We present \texttt{mcos-decoder}, a small
($\sim$\,270 lines, NumPy + SciPy only) pure-Python package focused on
extracting per-frame bounding-box labels from \texttt{groundTruth}
objects, complementing the broader \texttt{mat-io}~\cite{matio} parser
by exposing a single-purpose API
(\texttt{bboxes = load\_groundtruth(file)}). We validate the decoder on
the Halmstad Drone Detection Dataset (365 thermal IR videos, 4 classes),
recovering 102{,}476 bounding boxes that match the official MATLAB
reference for every sequence we tested. The package is released under
the MIT licence and is available on PyPI.
\end{abstract}

\keywords{MATLAB \and MCOS \and groundTruth \and Computer Vision Toolbox \and Python \and MAT-file format \and dataset tooling}

\section{Introduction}

The MATLAB Computer Vision Toolbox stores video annotations created by
the Video Labeler app as instances of the \texttt{groundTruth}
class~\cite{matlabcv}. These objects are saved into \texttt{.mat} files,
but unlike ordinary MATLAB structs they are serialised through MATLAB's
MCOS (\emph{MATLAB Class Object Storage}) mechanism, which writes the
data into an unnamed sub-file accessible only as
\texttt{\_\_function\_workspace\_\_}~\cite{scipyissue22736}. The
mainstream Python tools for reading \texttt{.mat} files
(\texttt{scipy.io}, \texttt{mat73}~\cite{mat73},
\texttt{pymatreader}~\cite{pymatreader}) treat these objects as opaque
blobs, so any researcher who wants to use a dataset distributed in this
form has historically needed a MATLAB licence.

The problem is documented in long-standing open issues: SciPy
\#22736~\cite{scipyissue22736} (filed 2025, still open at the time of
writing), the \texttt{DroneDetectionThesis} repository's
\#3~\cite{droneissue3} (filed February 2024, still open), and several
Stack Overflow threads~\cite{somat}. The C\#/.NET library
\texttt{MatFileHandler}~\cite{matfilehandler} provides a partial reader,
and its accompanying \texttt{Objects.md} notes are the canonical
reverse-engineering reference for the MCOS format. In Python, the
general-purpose \texttt{mat-io} parser~\cite{matio} (BSD-3-Clause, on
PyPI since April 2025) decodes MCOS opaque objects and exposes class
properties for downstream traversal.

This paper describes \texttt{mcos-decoder}, which occupies a
complementary niche to \texttt{mat-io}: a small, focused tool that
returns directly the per-frame bounding-box list expected by
computer-vision pipelines (e.g.\ tracker training, dataset loaders),
without requiring the caller to traverse MCOS class structures. The
implementation is short ($\sim$\,270 lines, NumPy + SciPy only),
permissively licensed under MIT, and validated against a published
365-video dataset.

\section{Background}

A MAT v5 file is a stream of typed elements: each element has an 8-byte
tag (type, size) followed by a payload aligned to 8 bytes. Type 14
(\texttt{miMATRIX}) elements contain nested element streams; numeric
types contain raw arrays~\cite{matformat}. SciPy's \texttt{loadmat}
parses ordinary numeric/struct/cell arrays but skips MCOS objects: when
it encounters one it returns a \texttt{MatlabOpaque} record whose four
fields (\texttt{s0}--\texttt{s2}, \texttt{arr}) describe only the
variable name, MCOS marker, and class name. The actual class data is
written to a separate top-level array which SciPy renames
\texttt{\_\_function\_workspace\_\_}.

The contents of \texttt{\_\_function\_workspace\_\_} are themselves a
nested MAT5 stream, but the structure depends on the MCOS class. For
\texttt{groundTruth}, empirical inspection across multiple sample files
yields the layout summarised in Table~\ref{tab:layout}.

\begin{table}[ht]
\centering
\caption{Empirical \texttt{groundTruth} layout in
\texttt{\_\_function\_workspace\_\_} (MATLAB R2020a--R2023a).}
\label{tab:layout}
\begin{tabular}{cll}
\toprule
Cell size (B) & Depth & Meaning \\
\midrule
80   & 5 & Per-frame bbox cell (4-element \texttt{double}). \\
112  & 5 & Per-frame empty marker (target absent). \\
72   & 5 & Per-frame duration / timestamp (skip). \\
\bottomrule
\end{tabular}
\end{table}

\section{Existing Readers and Positioning}

Table~\ref{tab:tools} summarises the Python and .NET tools that
interact with MAT-files containing MCOS objects, and locates
\texttt{mcos-decoder} within that ecosystem.

\begin{table}[ht]
\centering
\caption{Existing readers that touch MAT-files containing MCOS objects.
\emph{MCOS bytes} = the tool surfaces the opaque
\texttt{\_\_function\_workspace\_\_} buffer to the caller; \emph{MCOS
parsed} = it decodes class properties; \emph{Bboxes direct} = a single
call returns \texttt{groundTruth} per-frame boxes.}
\label{tab:tools}
\begin{tabular}{lccccc}
\toprule
Tool & Pure Python & MAT v5 & MCOS bytes & MCOS parsed & Bboxes direct \\
\midrule
\texttt{scipy.io.loadmat}                       & \checkmark & \checkmark & \checkmark & $\times$    & $\times$ \\
\texttt{mat73}~\cite{mat73}                     & \checkmark & v7.3 only  & $\times$   & $\times$    & $\times$ \\
\texttt{pymatreader}~\cite{pymatreader}         & \checkmark & \checkmark & \checkmark & $\times$    & $\times$ \\
\texttt{MatFileHandler}~\cite{matfilehandler}   & .NET only  & \checkmark & \checkmark & partial     & $\times$ \\
\texttt{mat-io}~\cite{matio}                    & \checkmark & \checkmark & \checkmark & \checkmark  & $\times$ \\
\midrule
\textbf{\texttt{mcos-decoder} (ours)} & \textbf{\checkmark} & \textbf{\checkmark} & \textbf{\checkmark} & \textbf{focused} & \textbf{\checkmark} \\
\bottomrule
\end{tabular}
\end{table}

\texttt{mcos-decoder} differs from \texttt{mat-io} in scope and intent:
where \texttt{mat-io} is a general MCOS-class parser that returns
property objects requiring further user traversal,
\texttt{mcos-decoder}'s entire public surface is a function that takes
a path and returns the per-frame bbox list expected by computer-vision
tooling. Its small footprint also makes it suitable for vendoring into
pipelines that cannot pull a heavier dependency, and the implementation
doubles as a compact pedagogical reference for the \texttt{groundTruth}
serialisation layout.

\section{Implementation}

The package is two small modules:

\begin{itemize}
  \item \texttt{mcos\_decoder.stream} --- generic walker that yields
        \texttt{MatElement} records (type, offset, payload, children)
        for the entire MAT5 element tree.
  \item \texttt{mcos\_decoder.groundtruth} --- high-level extractor
        that walks the tree, collects per-frame label cells from
        \texttt{depth\,$=$\,5}, and returns a list of
        \texttt{(x, y, w, h)} tuples or \texttt{None}.
\end{itemize}

The full pipeline from a MATLAB-saved file to a per-frame bbox list is
shown in Figure~\ref{fig:pipeline}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  node distance=0.55cm,
  every node/.style={font=\small},
  box/.style={rectangle, draw, rounded corners, minimum width=8.5cm,
              minimum height=0.7cm, align=center, inner sep=4pt},
  arrow/.style={-Stealth, thick}
]
  \node[box] (mat) {\texttt{labels.mat} (MAT v5 stream)};
  \node[box, below=of mat] (loadmat) {\texttt{scipy.io.loadmat}};
  \node[box, below=of loadmat] (fw)
    {opaque \texttt{\_\_function\_workspace\_\_} buffer};
  \node[box, below=of fw] (walk)
    {\texttt{MatStream.walk} (recursive element scan)};
  \node[box, below=of walk] (cells)
    {classify cells at depth 5 by byte size\\(80\,B bbox / 112\,B empty)};
  \node[box, below=of cells] (out)
    {\texttt{list[(x, y, w, h) | None]} per frame};
  \draw[arrow] (mat)     -- (loadmat);
  \draw[arrow] (loadmat) -- (fw);
  \draw[arrow] (fw)      -- (walk);
  \draw[arrow] (walk)    -- (cells);
  \draw[arrow] (cells)   -- (out);
\end{tikzpicture}
\caption{Decoder pipeline. \texttt{scipy.io.loadmat} surfaces the
opaque \texttt{\_\_function\_workspace\_\_} buffer;
\texttt{mcos-decoder} walks its MAT5 element tree, classifies cells at
depth 5 by byte size, and emits a per-frame bbox list.}
\label{fig:pipeline}
\end{figure}

The public API is one function:

\begin{lstlisting}
from mcos_decoder import load_groundtruth

bboxes = load_groundtruth("IR_DRONE_001_LABELS.mat")
# list[(x,y,w,h) | None] of length n_frames
\end{lstlisting}

A lower-level \texttt{MatStream} interface is also available for
callers that need to inspect a non-default MCOS layout (e.g.\ older
MATLAB versions or different toolbox classes).

\section{Validation}

We validated the decoder on the Halmstad Drone Detection
Dataset~\cite{svanstrom2021,halmstadzenodo}, which distributes 365
thermal IR videos of four target classes (\textsc{airplane},
\textsc{bird}, \textsc{drone}, \textsc{helicopter}). For every sequence
in Table~\ref{tab:halmstad-validation} we compared the decoder's
per-frame output to the video's frame count and the visible-frame count
obtainable from MATLAB's \texttt{selectLabels} on the original
\texttt{groundTruth}.

\begin{table}[ht]
\centering
\caption{Decoder validation on Halmstad IR sequences (one example per
class; all classes verified across the full 365-video set).}
\label{tab:halmstad-validation}
\begin{tabular}{lrrrc}
\toprule
Sequence & Frames & Decoded slots & With target & Match \\
\midrule
\texttt{IR\_AIRPLANE\_001}   & 322 & 322 & 322 (100\%) & \checkmark \\
\texttt{IR\_BIRD\_001}        & 310 & 310 & 235 (76\%)  & \checkmark \\
\texttt{IR\_DRONE\_001}       & 301 & 301 & 301 (100\%) & \checkmark \\
\texttt{IR\_HELICOPTER\_001}  & 301 & 301 & 301 (100\%) & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

The total decoded bounding-box volume for the full 365-video IR subset
is 102{,}476 boxes; spot checks against frame-by-frame visualisation
confirm that bbox positions and sizes round-trip correctly.

\section{Limitations and Future Work}

The current decoder addresses single-target-per-frame
\texttt{groundTruth} objects with the layout in Table~\ref{tab:layout}.
Multi-target labels (e.g.\ MOT-style annotations where one frame can
carry several boxes), categorical labels, polygon ROIs, and the related
\texttt{audioLabeler}/\texttt{signalLabeler} class families are planned
for v0.2; users requiring a richer view of MCOS class properties (and
not just per-frame boxes) are referred to \texttt{mat-io}~\cite{matio}.
Support for older MATLAB releases (pre R2020a) where the depth or cell
sizes differ slightly is straightforward through the existing
\texttt{frame\_depth}, \texttt{bbox\_size}, and \texttt{empty\_size}
parameters of \texttt{load\_groundtruth}.

\section{Availability and Reproducibility}

\texttt{mcos-decoder} is released under the MIT licence at
\href{https://github.com/bozdemir/mcos-decoder}{\texttt{github.com/bozdemir/mcos-decoder}}
and on PyPI as \texttt{mcos-decoder}. Contributions adding new dataset
layouts, additional MCOS classes, or pre-computed test fixtures are
welcome.

The validation in Table~\ref{tab:halmstad-validation} uses
\texttt{mcos-decoder} v0.1.0 (Python\,$\geq$\,3.9, NumPy\,$\geq$\,1.20,
SciPy\,$\geq$\,1.7). The Halmstad Drone Detection Dataset is publicly
archived at \href{https://doi.org/10.5281/zenodo.5500576}%
{\texttt{10.5281/zenodo.5500576}}~\cite{halmstadzenodo}. The
102{,}476-box total is reproduced by iterating the
\texttt{*\_LABELS.mat} files under the dataset's \texttt{IR/} root and
counting non-\texttt{None} returns:

\begin{lstlisting}
from pathlib import Path
from mcos_decoder import load_groundtruth

# IR/ is the unzipped Halmstad IR subset.
total = sum(b is not None
            for f in Path("IR").rglob("*_LABELS.mat")
            for b in load_groundtruth(f))
print(total)  # 102476
\end{lstlisting}

\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{matlabcv}
The MathWorks. \emph{Computer Vision Toolbox: groundTruth Class}.
\url{https://www.mathworks.com/help/vision/ref/groundtruth.html}.

\bibitem{matformat}
The MathWorks. \emph{MAT-File Format}.
\url{https://www.mathworks.com/help/pdf_doc/matlab/matfile_format.pdf}.

\bibitem{scipyissue22736}
SciPy contributors. \emph{ENH: io.loadmat: extract data from MATLAB
objects/classes}. GitHub Issue \#22736, opened 2025.
\url{https://github.com/scipy/scipy/issues/22736}.

\bibitem{droneissue3}
DroneDetectionThesis contributors. \emph{How to open bounding box
annotations in Python SciPy?} GitHub Issue \#3, opened February 2024.
\url{https://github.com/DroneDetectionThesis/Drone-detection-dataset/issues/3}.

\bibitem{somat}
Stack Overflow. \emph{Convert MATLAB groundTruth object to Python
readable format}. Multiple unanswered threads, 2022--2025.

\bibitem{matfilehandler}
A.~Mahalov. \emph{MatFileHandler: MATLAB .mat-file reading/writing API
for .NET Standard}.
\url{https://github.com/mahalex/MatFileHandler}.

\bibitem{matio}
\emph{mat-io: A pure-Python parser for MATLAB MAT-files with MCOS
class support}. PyPI package by GitHub user \texttt{foreverallama},
BSD-3-Clause, first released April 2025.
\url{https://pypi.org/project/mat-io/};
\url{https://github.com/foreverallama/matio}.

\bibitem{mat73}
\emph{mat73: Load MATLAB v7.3 .mat-files in Python}.
PyPI package, GPL-3.0.
\url{https://pypi.org/project/mat73/}.

\bibitem{pymatreader}
\emph{pymatreader: A Python module to read MATLAB-files}.
PyPI package, BSD-2-Clause.
\url{https://pypi.org/project/pymatreader/}.

\bibitem{svanstrom2021}
F.~Svanstr\"om, F.~Alonso-Fernandez, and C.~Englund.
\emph{A dataset for multi-sensor drone detection}.
Data in Brief, 39:107521, 2021.
\url{https://doi.org/10.1016/j.dib.2021.107521}.

\bibitem{halmstadzenodo}
F.~Svanstr\"om, F.~Alonso-Fernandez, and C.~Englund.
\emph{Drone detection dataset}. Zenodo, 2021.
\url{https://doi.org/10.5281/zenodo.5500576}.

\end{thebibliography}

\end{document}
